### 2020-10-27 submission/manuscript.tex
Edit Results section

\textbf{The impact of removing rare ASVs on the information represented
in each sample.} Removing rare ASVs will reduce the richness of ASVs and
proportionally increase the relative abundance of the remaining ASVs.
The result was expected to be a loss of information contained within
each sample. To quantify the effect of removing rare ASVs on the
information contained within each sample, I varied the minimum abundance
threshold to simulate removing ASVs of varying rarity from each sample.
The richness of ASVs in each sample (i.e.~the number of ASVs) decreased
by between 34.4 and 86.2\% when removing those ASVs that only appeared
once and by between 76.0 and 95.6\% when removing those that appeared
ten or fewer times from each sample (Figure 2A). Similarly, the Shannon
diversity decreased by between 1.8 and 15.9\% when removing ASVs that
only appeared once and by between 76.0 and 95.6\% when removing ASVs
that appeared ten or fewer times from each sample (Figure 2B). Next, I
assigned the ASVs to OTUs to assess the impact of removing rare ASVs on
higher level taxonomic groupings that are commonly used in microbial
ecology studies. Although pooling similar ASVs into OTUs reduced the
impact of removing the rare ASVs relative to the ASV-based analysis, the
minimum abundance threshold still decreased the richness of OTUs and the
diversity decreased relative to the full community (\textbf{Figure
S2AB}). In contrast to the richness and diversity measurements, the
Kullback--Leibler divergence compares the relative abundance of specific
ASVs or OTUs between representations of the community. I calculated the
Kullback--Leibler divergence from the full to pruned communities when
rare ASVs were removed. As the threshold for removing ASVs increased,
the amount of information lost also increased for both ASVs and OTUs
(Figure 2C and Figure S2C). The relative loss of information was
generally lower for OTUs than than it was for ASVs. Removing rare ASVs,
regardless of abundance threshold, had profound impacts on the
representation of the communities.

\textbf{Removing treatment group effects from community data.} Because
treatment effects often affect a sample's diversity and inter-sample
variation, I generated null distributions for each study by randomizing
the number of times each ASV was observed in each sample such that the
total number of sequences in each sample and the total number of times
each ASV was observed across all samples in the study was the same as
was originally observed. This effectively made every community in a
study a statistical sample of the study-wide composite community
distribution. For example, after this procedure, the 490 samples from
the human dataset would be expected to have the same richness and
diversity of ASVs and one would not expect to find treatment-based
effects between the samples. Because of the risk of bias if only one
representation of the null distribution was generated, I generated 100
randomized datasets for each study. The trends between removing rare
ASVs and the richness, diversity, and information loss that were
identified using the observed community community distribution data were
also identified with the data from the null distribution; however, the
losses were larger when using the null distribution data (\textbf{Figure
S3}). The null distribution data were used in the remainder of the study
to minimize the risk of bias.

\textbf{The impact of removing rare ASVs on the information represented
between samples.} Considering the loss of richness, diversity, and
information when a community has its rarest ASVs removed, it seemed
likely that the relationship between communities would also be altered.
To assess the impact of removing rare ASVs on measures of alpha
diversity between samples I calculated the coefficients of variation
(COVs, i.e.~standard deviation divided by the mean) for richness and
diversity for each study at multiple abundance thresholds. The COVs for
richness of ASVs across the studies after removing singletons were
between 3.6 and 32.7-times larger than they were without removing
singleton ASVs (Figure 3A). Similarly, the COVs for the diversity of
ASVs were between 1.8 and 20.4-times larger when singletons were removed
than when they were not removed (Figure 3B). To assess the impact of
removing rare ASVs on measures of beta diversity between samples I
calculated the COVs of the Bray-Curtis distances between samples within
the same study at multiple abundance thresholds. The COVs between
Bray-Curtis distances within a study when singletons were removed was
between 1.3 and 18.6-times larger than when they were not removed
(Figure 3C). Increasing the minimum abundance threshold increased the
COVs between samples when using metrics of alpha and beta diversity.
When ASVs were clustered into OTUs the difference in COVs was less than
it was for the ASVs (Figure S4). These results indicate that removing
rare ASVs increases the dissimilarity between samples, which could have
a significant impact on the statistical power to detect differences
between treatment groups.

\textbf{The impact of removing rare ASVs on the ability to detect
statistically significant differences between treatment groups.} To test
the effect of increased inter-sample variation, I randomly assigned
samples to one of two treatment groups. In the first treatment group,
communities were randomly sampled from the null distribution as
described above. For the second treatment group, I increased the
abundance of 10\% of the ASVs in the pooled study distribution by 5\%. I
randomly generated 100 simulated sets of treatment groups and samples. I
then tested the ability to detect a difference between the two treatment
groups using alpha and beta diversity metrics. The fraction of
significant tests was a measurement of the statistical power to detect
the difference between the treatment groups. When considering the
differences in richness and diversity, the marine dataset yielded no
simulated sets that were statistically significant, which was likely due
to the small number of samples in the study (N=7). Among the remaining
datasets, the power to detect a difference in the richness of ASVs
ranged between 0.10 and 0.49 and between 0.10 and 0.53 to detect a
difference in diversity when using a Wilcox test (Figure 4A). When
singleton ASVs were removed, the power to detect a difference in the
diversity of ASVs dropped by between 27.3 and 92.9\% and by between 40.0
and 93.3\% (Figure 4B). The effect of removing rare ASVs on the richness
of OTUs and their diversity was similar (Figure S5AB). I used the
Bray-Curtis dissimilarity index to compare the simulated communities
within each dataset and calculated the power to detect differences
between the two simulated treatment groups using the analysis of
molecular variance (also called PERMANOVA) (Figure 4C and S5C). Without
removing rare sequences, the power to detect a difference between the
two simulated treatment groups varied between 0.41 and 1.00. Aside from
3 datasets, the power to detect differences dropped by between 6.5 and
64.0\% when singletons were removed. However, when ASVs that occurred 10
or fewer times were removed from each sample, the power to detect
differences dropped by 12.0 and 97.2\%; similar results were observed
when ASVs were clustered into OTUs. Removing rare ASVs reduced the
ability to detect simulated treatment effects using metrics commonly
used to compare microbial communities.

\textbf{The impact of removing rare ASVs on the probability of falsely
detecting a difference between treatment groups.} Observing reduced
ability to detect differences between communities when rare ASVs were
removed from each sample, I next asked whether removing rare ASVs could
lead to falsely claiming that a treatment effect had a significant
effect on community diversity and structure. First, I sampled sequences
from the null distribution for each dataset and randomly assigned each
sample to one of two treatment groups and determined the richness and
diversity of ASVs and OTUs. Testing at an experiment-wise error rate of
0.05, I expected 5\% of the iterations for each dataset to yield a
significant test result. Indeed, there was no evidence that removing
rare ASVs resulted in an inflated experiment-wise error rate. The
average fraction of significant tests did not meaningfully vary from
0.05 across the minimum abundance threshold, dataset, metric of
describing sample alpha-diversity, or whether the abundance of ASVs or
OTUs were used (Figure 5A and S6A). Similarly, the average fraction of
significant tests did not meaningfully vary from 0.05 when using
analysis of molecular variance to compare communities using Bray-Curtis
distances (Figure 5A and S6A). Second, I again sampled sequences from
the null distribution, but assigned samples to one of two treatment
groups based on the number of sequences in each sample. The samples with
fewer than the median number of sequences for the dataset were assigned
to one group and those with more than the median were assigned to the
other. This exaggerated bias has been observed in comparisons of the
lung and oral microbiota because of the larger number of non-specific
amplicons that can be sequenced from lung samples relative to those in
the oral cavity leading to a significant difference in sequencing depth
between treatment groups {[}\textbf{REF}{]}. When rare sequences were
not removed, the fraction of significant tests did not differ from 5\%
for comparing the richness, their diversity, or Bray-Curtis distances
(Figure 5B and S6B). However, when rare taxa of any frequency were
removed, the probability of falsely detecing a difference as signifiant
increased with the definition of rarity (Figure 5B and S6B). Not
including the small marine dataset, the average fraction the average
fraction of falsely detecting a difference across datasets when only
singletons were removed was 92.45\%. If there is any relationship
between the number of sequences and the treatment group, the risk of
falsely rejecting the null hypothesis is inflated when researchers use
the strategy of removing rare sequences. The most conservative approach
is to not remove low abundance sequences.

\textbf{\emph{Conclusion.}} Removing rare sequences decreases the
diversity represented by 16S rRNA gene sequence data and increases the
variation between samples. Such impacts will hinder the statistical
power to differentiate between treatment groups. Instead of removing
rare sequences, researchers should focus on optimizing their sequence
generation to minimize the amount of PCR and sequencing errors. In
addition, samples should be rarefied to a common number of sequences
across samples without prior culling of rare sequences. The number of
artifacts is correlated to the number of sequences being considered.
With this in mind, rarefaction allows one to control for uneven sampling
effort and to control for the number of artifacts in the analysis.

@@

\begin{tabular}{lrrrrr}
\toprule
\textbf{Study (Ref)} & \textbf{Samples} & \textbf{\makecell[c]{Total\\sequences}} & \textbf{\makecell[c]{Median\\sequences}} & \textbf{\makecell[c]{Range of\\sequences}} & \textbf{\makecell[c]{Fold-difference\\between largest\\and smallest sample}}\\
\midrule
Bioethanol (NA) & 95 & 3,972,943 & 16,015 & 3,688-356,136 & 96.6\\
Human (NA) & 490 & 20,909,768 & 32,505 & 10,523-430,415 & 40.9\\
Lake (NA) & 52 & 3,169,868 & 69,041 & 15,347-112,871 & 7.4\\
Marine (NA) & 7 & 1,391,396 & 193,464 & 133,516-254,060 & 1.9\\
Mice (NA) & 348 & 2,813,747 & 6,477 & 1,804-30,565 & 16.9\\
Peromyscus (NA) & 111 & 1,555,545 & 12,446 & 4,464-33,644 & 7.5\\
Rainforest (NA) & 69 & 946,295 & 11,561 & 4,932-37,767 & 7.7\\
Rice (NA) & 490 & 22,591,168 & 43,216 & 2,776-193,464 & 69.7\\
Seagrass (NA) & 286 & 4,130,454 & 13,567 & 1,803-45,191 & 25.1\\
Sediment (NA) & 58 & 1,154,174 & 17,584 & 7,685-68,321 & 8.9\\
Soil (NA) & 18 & 956,656 & 51,844 & 47,806-59,956 & 1.3\\
Stream (NA) & 201 & 21,162,574 & 90,159 & 9,175-390,964 & 42.6\\
\bottomrule
\end{tabular}

@@

\textbf{Figure 2.} v

>>>>

\textbf{The impact of removing rare ASVs on the information represented
in each sample.} Removing rare ASVs will reduce the richness of ASVs
(i.e.~the number of ASVs per sample) and increase the relative abundance
of the remaining ASVs. To quantify the effect of removing rare ASVs on
the information contained within each sample, I varied the minimum
abundance threshold to simulate removing ASVs of varying rarity from
each sample. The richness of ASVs in each sample decreased by between
34.4 and 86.2\% (peromyscus and soil) when removing those ASVs that only
appeared once and by between 76.0 and 95.6\% (sediment and soil) when
removing those that appeared ten or fewer times from each sample (Figure
2A). Similarly, the Shannon diversity decreased by between 1.8 and
15.9\% (human and soil) when removing ASVs that only appeared once and
by between 5.4 and 35.4\% (human and seagrass) when removing ASVs that
appeared ten or fewer times from each sample (Figure 2B). Next, I
assigned the ASVs to OTUs, which were defined as a group of ASVs that
were more than 97\% similar to each other to assess the impact of
removing rare ASVs on higher level taxonomic groupings that are commonly
used in microbial ecology studies. Although pooling similar ASVs into
OTUs reduced the impact of removing the rare ASVs relative to the
ASV-based analysis, the minimum abundance threshold still decreased the
richness of OTUs and the diversity decreased relative to the full
community (Figure S2AB). In contrast to the richness and diversity
measurements, the Kullback--Leibler divergence compares the relative
abundance of specific ASVs or OTUs between representations of the
community. I calculated the Kullback--Leibler divergence between the
full communities and those where rare ASVs were removed. As the
threshold for removing ASVs increased, the amount of information lost
also increased for both ASVs and OTUs (Figure 2C and Figure S2C). The
relative loss of information was generally smaller for OTUs than than it
was for ASVs. Removing rare ASVs, regardless of abundance threshold, had
profound impacts on the representation of the communities.

\textbf{Removing treatment group effects from community data.} Because
treatment effects often affect a sample's diversity and inter-sample
variation, I generated null distributions for each study by randomizing,
without replacement, the number of times each ASV was observed in each
sample such that the total number of sequences in each sample and the
total number of times each ASV was observed across all samples in the
study was the same as was originally observed. This effectively made
every community in a study a statistical sample of the study-wide
composite community distribution. For example, after this procedure, the
490 samples from the human dataset would be expected to have the same
richness and diversity of ASVs and one would not expect to find
treatment-based effects between the samples. Because of the risk of bias
if only one representation of the null distribution was generated, I
generated 100 randomized datasets for each study. The trends between
removing rare ASVs and the richness, diversity, and information loss
that were identified using the observed community community distribution
data were also identified with the data from the null distribution;
however, the losses were larger when using the null distribution data
(Figure S3). The null distribution data were used in the remainder of
the study to minimize the risk of bias.

\textbf{The impact of removing rare ASVs on the information represented
between samples.} Considering the loss of richness, diversity, and
information when a community has its rarest ASVs removed, it seemed
likely that the relationship between communities would also be altered.
To assess the impact of removing rare ASVs on measures of alpha
diversity between samples I calculated the coefficients of variation
(COVs, i.e.~the standard deviation divided by the mean) for richness and
diversity for each study at multiple abundance thresholds. The COVs for
the richness of ASVs across the studies after removing singletons were
between 3.6 and 32.7-times larger than they were without removing
singleton ASVs (mice and stream; Figure 3A). Similarly, the COVs for the
diversity of ASVs were between 1.8 and 20.4-times larger when singletons
were removed than when they were not removed (mice and rice; Figure 3B).
To assess the impact of removing rare ASVs on measures of beta diversity
between samples, I calculated the COVs of the Bray-Curtis distances
between samples within the same study at multiple abundance thresholds.
The COVs between Bray-Curtis distances within a study when singletons
were removed was between 1.3 and 18.6-times larger than when they were
not removed (mice and stream; Figure 3C). When ASVs were clustered into
OTUs the difference in COVs was less than it was for the ASVs (Figure
S4). These results indicate that removing rare ASVs increases the
dissimilarity between samples, which could have a significant impact on
the statistical power to detect differences between treatment groups.

\textbf{The impact of removing rare ASVs on the ability to detect
statistically significant differences between treatment groups.} To test
the effect of increased inter-sample variation, I randomly assigned
samples to one of two treatment groups. In the first treatment group,
communities were randomly sampled from the null distribution as
described above. For the second treatment group, I randomly selected
10\% of the ASVs in the pooled study distribution to increase their
abundance by 5\%. I randomly generated 100 simulated sets of treatment
groups and samples. I then tested the ability to detect a difference
between the two treatment groups using alpha and beta diversity metrics.
The fraction of significant tests was a measurement of the statistical
power to detect the difference between the treatment groups. When
considering the differences in richness and diversity, the marine
dataset yielded no simulated sets that were statistically significant,
which was likely due to the small number of samples in the study (N=7).
Among the remaining datasets, the power to detect a difference in the
richness of ASVs ranged between 0.10 and 0.49 (sediment and stream) and
between 0.10 and 0.53 (rainforest and stream) to detect a difference in
diversity when using a Wilcox test (Figure 4A). When singleton ASVs were
removed, the power to detect a difference in the diversity of ASVs
dropped by between 27.3 and 92.9\% (bioethanol and soil) and by between
40.0 and 93.3\% (rainforest and soil; Figure 4B). The effect of removing
rare ASVs on the richness of OTUs and their diversity was similar
(Figure S5AB). I used the Bray-Curtis dissimilarity index to compare the
simulated communities within each dataset and calculated the power to
detect differences between the two simulated treatment groups using the
analysis of molecular variance (also called PERMANOVA) (Figure 4C and
S5C). Without removing rare sequences, the power to detect a difference
between the two simulated treatment groups varied between 0.41 and 1.00
(rainforest and rainforest). Aside from the bioethanol, human, and mice
datasets, the power to detect differences dropped by between 6.5 and
64.0\% (soil and rice) when singletons were removed. However, when ASVs
that occurred 10 or fewer times were removed from each sample, the power
to detect differences dropped by 12.0 and 97.2\% (human and peromyscus);
similar results were observed when ASVs were clustered into OTUs.
Removing rare ASVs reduced the ability to detect simulated treatment
effects using metrics commonly used to compare microbial communities.

\textbf{The impact of removing rare ASVs on the probability of falsely
detecting a difference between treatment groups.} I next asked whether
removing rare ASVs could lead to falsely claiming that a treatment
effect had a significant effect on community diversity and structure.
First, I sampled sequences from the null distribution for each dataset
and randomly assigned each sample to one of two treatment groups and
determined the richness and diversity of ASVs and OTUs. Testing at an
experiment-wise error rate of 0.05, I expected 5\% of the iterations for
each dataset to yield a significant test result. Indeed, there was no
evidence that removing rare ASVs resulted in an inflated experiment-wise
error rate. The average fraction of significant tests did not
meaningfully vary from 0.05 across the minimum abundance threshold,
dataset, metric of describing sample alpha-diversity, or whether the
abundance of ASVs or OTUs were used (Figure 5A and S6A). Similarly, the
average fraction of significant tests did not meaningfully vary from
0.05 when using analysis of molecular variance to compare communities
using Bray-Curtis distances (Figure 5A and S6A). Second, I again sampled
sequences from the null distribution, but assigned samples to one of two
treatment groups based on the number of sequences in each sample. The
samples with fewer than the median number of sequences for the dataset
were assigned to one group and those with more than the median were
assigned to the other. This exaggerated bias has been observed in
comparisons of the lung and oral microbiota because of the larger number
of non-specific amplicons that can be sequenced from lung samples
relative to those in the oral cavity leading to a significant difference
in sequencing depth between treatment groups {[}{]}. When rare sequences
were not removed, the fraction of significant tests did not differ from
5\% for comparing the richness, their diversity, or Bray-Curtis
distances (Figure 5B and S6B). However, when rare taxa of any frequency
were removed, the probability of falsely detecing a difference as
signifiant increased with the definition of rarity (Figure 5B and S6B).
Not including the small marine dataset, the average fraction the average
fraction of falsely detecting a difference across datasets when only
singletons were removed was 92.45\%. If there is any relationship
between the number of sequences and the treatment group, the risk of
falsely rejecting the null hypothesis is inflated when researchers use
the strategy of removing rare sequences. The most conservative approach
is to not remove low abundance sequences.

\subsection{Discussion}\label{discussion}

Removing rare sequences decreases the diversity represented by 16S rRNA
gene sequence data and increases the variation between samples. Such
impacts will hinder the statistical power to differentiate between
treatment groups. Instead of removing rare sequences, researchers should
focus on optimizing their sequence generation to minimize the amount of
PCR and sequencing errors. In addition, samples should be rarefied to a
common number of sequences across samples without prior culling of rare
sequences. The number of artifacts is correlated to the number of
sequences being considered. With this in mind, rarefaction allows one to
control for uneven sampling effort and to control for the number of
artifacts in the analysis.

@@

\begin{tabular}{lrrrrr}
\toprule
\textbf{Study} & \textbf{Samples} & \textbf{\makecell[c]{Total\\sequences}} & \textbf{\makecell[c]{Median\\sequences}} & \textbf{\makecell[c]{Range of\\sequences}} & \textbf{\makecell[c]{Fold-difference\\between largest\\and smallest sample}}\\
\midrule
Bioethanol & 95 & 3,972,943 & 16,015 & 3,688-356,136 & 96.6\\
Human & 490 & 20,909,768 & 32,505 & 10,523-430,415 & 40.9\\
Lake & 52 & 3,169,868 & 69,041 & 15,347-112,871 & 7.4\\
Marine & 7 & 1,391,396 & 193,464 & 133,516-254,060 & 1.9\\
Mice & 348 & 2,813,747 & 6,477 & 1,804-30,565 & 16.9\\
Peromyscus & 111 & 1,555,545 & 12,446 & 4,464-33,644 & 7.5\\
Rainforest & 69 & 946,295 & 11,561 & 4,932-37,767 & 7.7\\
Rice & 490 & 22,591,168 & 43,216 & 2,776-193,464 & 69.7\\
Seagrass & 286 & 4,130,454 & 13,567 & 1,803-45,191 & 25.1\\
Sediment & 58 & 1,154,174 & 17,584 & 7,685-68,321 & 8.9\\
Soil & 18 & 956,656 & 51,844 & 47,806-59,956 & 1.3\\
Stream & 201 & 21,162,574 & 90,159 & 9,175-390,964 & 42.6\\
\bottomrule
\end{tabular}

@@

\textbf{Figure 2.}
